### 2017-04-20 Manuscripts/cog-psych/cogpsych.tex
edits through results

As noted above, the constraint that new concepts should differ from what is already known has been largely overlooked in previous work. This is no doubt in part due to the vague definition of what it means for a concept to be ``different'': A generated category may be different from what is already known in any number of respects. Towards providing a more precise definition of the role of contrast in creative generation, we formalized contrast in a novel exemplar model, PACKER  ({\em Producing Alike and Contrasting Knowledge using Exemplar Representations}). PACKER explains category generation as a balance between two fundamental constraints: generated categories should not be similar to known categories, and exemplars within each category should be similar to one another. These ideas are implemented within the well-studied exemplar framework: the PACKER model is an extension of the influential Generalized Context Model of category learning \citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}.

@@

PACKER sets $f(x_j)$ depending on exemplar $x_j$'s category membership: $f(x_j) = \gamma$ if $x_j$ is a member of the target category, and $f(x_j) = \gamma - 1$ if $x_j$ is a member of a contrast category. $\gamma$ is thus a free parameter ($0 \leq \gamma \leq 1$) controlling the trade-off between within- and between-class similarity. For example, when $\gamma = 0.5$, $f(x_j) = 0.5$ for members of the target category and $f(x_j) = -0.5$ for members of other categories; thus, the model is likely to generate items that are similar to members of the target category but are not similar to members of other categories. In this way, $\gamma = 1$ produces exclusive consideration of target-category members, and $\gamma = 0$ produces exclusive consideration of opposite-category members. The $\gamma$ parameter thus specifies a wide breadth of possible approaches; by fitting it to a dataset, one can describe the relative roles of between-class contrast and within-class similarity in generation. See Figure \ref{fig:packer-examples} for an illustration of how $\gamma$ controls the relative influence of within category similarity and contrast to other categories.

@@

It is also worth noting that PACKER represents just one of many possible models that incorporate contrast in category generation. For example, although it may be possible to extend \cite{jern2013probabilistic}'s hierarchical sampling model to include contrast, this mechanism would be at odds with a conjugate hierarchical Bayesian framework. In contrast, these ideas emerge naturally from the exemplar view: we have not modified any of the core elements of the GCM in defining the PACKER model, we simply aggregated similarity slightly differently. Thus, beyond formally specifying the role of contrast in creative generation, the PACKER account allows us to evaluate the well-understood principles of the exemplar view within the comparatively understudied field of conceptual generation. Nonetheless, we will discuss integrating our approaches in the General Discussion.

@@

\jlanote{I forget, but did JK13 explicitly have in their instructions that the new category be different? If so, we should mention the issue there as another problem with their study for us.}
\section{Experiment 1}

To begin our investigation, we sought to extend the early evidence obtained from our analysis of the \cite{jern2013probabilistic} data, under a variety of learning conditions and using more standard stimulus materials. We used an artificial stimulus design, two dimensional domain of squares, varying in color and size (see Figure \ref{fig:e1-conditions}, panel A). These dimensions have been used in numerous classification learning studies \cite[e.g.,][]{conaway2016similar,conaway2016generalization,shepard1961learning,nosofsky1994comparing}. Unlike those used in the \cite{jern2013probabilistic} experiments, proximity along these dimensions aligns more directly with perceptual similarity, allowing us to evaluate the role of category contrast in creative generation more precisely. To extend the evidence provided by the \cite{jern2013probabilistic} data, we tested the effects of category contrast after learning one category from a set of category structures. We did so by training participants on categories possessing qualitatively different distributional structures, as shown in Figure \ref{fig:e1-conditions}.

@@

In this way, we can analyze the results of the first experiment as a conceptual replication of the classic finding that generated categories tend to share distributional properties with known categories in the domain \citep[see][]{jern2013probabilistic,ward1994structured}. From these results, we can predict that, in each condition, participants should generate categories that are distributionally similar to the experimenter-defined category: In the Cluster condition, generated categories should be tightly clustered. In the Row condition, generated categories should very more along the X-axis than the Y-axis. In XOR condition, generated categories should be widely distributed across both dimensions, and the two dimensions should be positively correlated.

Interestingly, the XOR condition also offers a dissociation between the roles of category contrast and the emulation of distributional structure: widely-distributed, positively-correlated categories would need to lie along the positive diagonal of the space (that is the only place they ``fit''), which is already occupied by the experimenter-defined category. Thus, if contrast plays a role, exemplars in the generated categories of participants in the XOR condition may not be positively correlated--they may be negatively correlated instead.

@@

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{e1-distanceplots.pgf}
    \caption{Experiment 1 results. {\em Left}: Frequency of generation as a function of distance from members of the experimenter-defined category. {\em Right}: Scatter plot of within-class versus between-class distance in each of the participant-generated categories.}

    \nbcnote{Need to check on how nice the journal is about color figs!}
	\jlanote{Is there something systematic about XOR for how the generated category varies for smaller/larger within-category distances. Also, should switch "class" to category here.}
    \label{fig:e1-distanceplots}
    \end{center}
\end{figure}

Figure \ref{fig:e1-distanceplots} also depicts, for each participant, the average distance of members within the generated category ({\em within-class} distance) against the average distance between members of the generated and experimenter-defined category ({\em between-class} distance). These data also reveal a systematic pattern: the majority of participants generated categories with more between-class distance than within-class distance. That is, members of the generated category tended to be more similar to one another than to members of the experimenter-defined category. To formally evaluate this claim, we conducted t-tests comparing the amount of within- and between- class distance in each condition. All conditions possessed greater between-class distance: Cluster, $t(63) = 11.43$, $p < 0.001$; Row, $t(60) = 13.16$, $p < 0.001$; and XOR, $t(57) = 3.64$, $p < 0.001$. These results provide clear evidence of an effect of category contrast: participants prefer to generate categories that are dissimilar to the learned category but maintain some level of internal cohesion.

@@

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{e2-distanceplots.pgf}
    \caption{Experiment 2 results. {\em Left}: Frequency of generation as a function of distance from members of the experimenter-defined category. {\em Right}: Scatter plot of within-class versus between-class distance in each of the participant-generated categories.}
    \label{fig:e2-distanceplots}
    \end{center}
\end{figure}

We began our analysis by again testing for the broad influence of category contrast on generation. As in Experiment 1, we computed the frequency each stimulus was generated as a function of its average distance from members of the experimenter defined category, as well as each participant's average within- and between- category distance. These data, shown in Figure \ref{fig:e2-distanceplots}, yield highly similar results. Stimuli that are more distant from members of the experimenter-defined category were more frequently generated, and the categories in each condition tended to possess more between-class than within-class distance: Bottom, $t(60) = 5.5$, $p < 0.001$; Middle, $t(60) = 2.71$, $p = 0.009$.

We did, however, observe a notable subgroup of participants in each condition who generated categories with more within-class than between-class distance. Upon manual inspection, many of these individuals appear to have assumed a `Corners' strategy, placing exemplars in disparate corners of the space, thus producing much more within-class distance, see Figure \ref{fig:e2-samples} for examples.

@@

In Experiment 2, we replicated the core findings from Experiment 1: stimuli are more likely to be generated if they are distant from exemplars in other categories, and most participants generate categories with more between-class than within-class distance. However, we additionally found that the {\em position} of a previously learned category (rather than its distributional structure) influences the types of categories people generate: participants who learned the `Middle' type were more likely to generate categories spanning the entire Y-axis of the space. Participants who learned the `Bottom' type were less likely to do so as a result of the presence of opposite category exemplars in the lower regions of the space.

@@

Experiments 1 and 2 revealed systematic and strong effects of category contrast on creative generation. In this section, we report the results of simulations with formal models aimed at explaining our observations. Specifically, we present simulations from the PACKER model, as well as a `copy-and-tweak' model (discussed in Section \ref{section:copytweak-packer}), defined as a variant of PACKER with the $\gamma$ parameter constrained to be one. The comparison of these two models serves to highlight the explanatory role of contrast within PACKER's framework: if contrast affords little explanatory advantage, then the two accounts should produce an equally strong account. We also present simulations from an implementation of the hierarchical sampling model proposed by \cite{jern2013probabilistic}, described in-depth in Appendix \ref{ap:hsampling-definition}. The comparison between the hierarchical sampling model and PACKER is meant to emphasize the necessity of contrast and demonstrate it cannot be explained by emulating of distributional structure: whereas PACKER is insensitive to the distributional structure of learned categories (relying only on within- and between-class similarity), the hierarchical sampling model generates categories exclusively on the basis of knowledge of how existing classes are distributed. Our approach in this section is to first broadly evaluate and compare the quality of each model's account to our entire dataset (Experiments 1 and 2 combined), then we more specifically describe the strengths and weakness of each model's account.

@@

Through comparison with the copy-and-tweak model, Figure \ref{fig:packer-loglike} more clearly demonstrates the robustness of the explanatory gains yielded by PACKER's category contrast mechanism. Plotted is PACKER's log-likelihood as a function of $\gamma$ parameter. The model's other parameters ($c$, $\theta$) were set according to copy-and-tweak's best fits from Table \ref{table:global-model-fits}, and thus when $\gamma=1$, the models are equivalent. The figure clearly shows a ``sweet spot'':  a convex region in which PACKER achieves superior fits as a result of changes to $\gamma$. The best fitting values lie well below the value of 1 assumed by the copy-and-tweak model, which demonstrates the robustness of the contrast effect (though note PACKER achieves even better fits when its other parameters are fitted, as in Table \ref{table:global-model-fits}). Notably, however, the copy-and-tweak parameterization ($\gamma=1$) performs better than the `contrast-only' parameterization of the model ($\gamma = 0$). Thus, it is evident that the data is better explained when both within-class similarity and category contrast is considered.

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{packer-loglike.pgf}
    \caption{PACKER's fit as a function of its prioritization of within-class and between-class similarity (using the $\gamma$ parameter). To facilitate comparison, PACKER's other parameters ($c$, $\theta$) were set to the best fitting values obtained for copy-and-tweak in Table \ref{table:global-model-fits}. }
    \label{fig:packer-loglike}
    \end{center}
\end{figure}

@@

In the framework proposed by PACKER, the cluster and corners profiles arise based on different prioritization of within-class similarity versus between-class contrast, and the row and column profiles arise based on the prioritization of each dimension in the computation of similarity. For example, in the cluster profile, there is a high degree of within-class similarity along both dimensions, whereas in the corners profile there is minimal within-class similarity. Thus, PACKER's proposal is that these individual differences arise as a result of different priorities: While the tight cluster configuration can be considered PACKER's `default' mode (as it maximizes within-class similarity), the corners profile can be produced when between-class contrast is put at a higher priority (i.e., $\gamma = 0$).

Likewise, in the row and column profiles, there is a high degree of within-class similarity along one dimension but not the other. These differences likely arise due to a differential focus on one dimension over another, and thus they can be produced by changes to PACKER's attention weights, $w$ (see Equation \ref{eq:similarity}). Traditionally, the attention weights in exemplar models are thought to reflect the diagnostic value of each dimension towards classifying the known category members \citep{nosofsky1984choice,nosofsky1986attention,kruschke1992alcove}, but within a generation context the weights specify the importance of within- and between-class similarity along each dimension. For example, if 100\% of attention is allocated along the X-axis, similarity along the Y-axis no longer influences performance: as a result, PACKER will create categories that are more widely distributed along the Y-axis, as similarity is not counted along that dimension. As a general principle, differentially weighting one dimension will result in the generation of categories that are more widely distributed along the ignored dimension, conforming to a row- or column-like arrangement. See Figure \ref{fig:packer-attention} for a depiction of how attention influences PACKER's performance.

@@

These patterns of performance clearly depict the interdependence between the distributional structure and location of generated concepts. Our results can, in some sense, be interpreted in terms of local minimization of between-class similarity: by distributing the generated category away from members of the experimenter-defined category, participants may increase the degree of between-class distance without drastically altering the degree of within-class similarity.

@@

The creative generation of novel concepts is an intriguing yet understudied topic in cognitive science. While the bulk of prior research on the topic has focused on the classic finding that generated concepts tend to be distributionally similar to known concepts, there has been little work addressing the role of contrast in creative generation: how is it that people are able to create something {\em different} from what is already known? This issue is of fundamental importance: in order to successfully generate something new, it must be different from what is known. We developed a novel, exemplar-based model, PACKER, which formally specifies the role of contrast in generation. The model proposes that categories are represented as exemplars in a multidimensional psychological space, and generation is constrained both by within-class and between-class similarity: exemplars belonging to the same category should be similar to one another, and exemplars belonging to different categories should not be similar to one another.

@@

The success of PACKER yields reveals keys insights about the nature of creative generation. First, by measuring PACKER's fit as a function of its prioritization of within- and between-class similarity, we observed that considering either constraint exclusively results in a relatively low-quality account. Instead, PACKER's best results were obtained when both constraints are considered, indicating that human learners do not generate novel concepts exclusively on the basis of within-class similarity or between class-contrast. This finding mirrors our behavioral results: it is clear that both constraints influence creative generation.

This finding is extended by our observation of the broad interdependence between the distributional structure (feature variance, correlation) and physical instantiation (location within the domain space) of generated categories. In Experiment 2, we found that the unoccupied regions of the domain influenced the distributional structure of categories, and in Experiments 1 and 2 we observed that generated categories assume different distributional structures depending on how they contrast with known categories. This interdependence is straightforwardly explained under PACKER's dual-constraint proposal: after first establishing a location for their generated category, participants distribute exemplars in order to minimize between-class similarity while not drastically decreasing within-class similarity.

>>>>

As noted above, the constraint that new concepts should differ from what is already known has been largely overlooked in previous work. This is no doubt in part due to the vague definition of what it means for a concept to be ``different'': A generated category may be different from what is already known in any number of respects. Towards providing a more precise definition of the role of contrast in creative generation, we formalized contrast in a novel exemplar model, PACKER  ({\em Producing Alike and Contrasting Knowledge using Exemplar Representations}). PACKER explains category generation as a balance between two fundamental constraints: generated categories should not be similar to known categories, and exemplars within each category should be similar to one another. These ideas are implemented within the well-studied exemplar framework -- the PACKER model is an extension of the influential Generalized Context Model of category learning \citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}.

@@

PACKER sets $f(x_j)$ depending on exemplar $x_j$'s category membership: $f(x_j) = \gamma$ if $x_j$ is a member of the target category, and $f(x_j) = \gamma - 1$ if $x_j$ is a member of a contrast category. $\gamma$ is thus a free parameter ($0 \leq \gamma \leq 1$) controlling the trade-off between within- and between-category similarity. For example, when $\gamma = 0.5$, $f(x_j) = 0.5$ for members of the target category and $f(x_j) = -0.5$ for members of other categories; thus, the model is likely to generate items that are similar to members of the target category but are not similar to members of other categories. In this way, $\gamma = 1$ produces exclusive consideration of target-category members, and $\gamma = 0$ produces exclusive consideration of opposite-category members. The $\gamma$ parameter thus specifies a wide breadth of possible approaches; by fitting it to a dataset, one can describe the relative roles of between-category contrast and within-category similarity in generation. See Figure \ref{fig:packer-examples} for an illustration of how $\gamma$ controls the relative influence of within category similarity and contrast to other categories.

@@

It is also worth noting that PACKER represents just one of many possible models that incorporate contrast in category generation. For example, although it may be possible to extend \cite{jern2013probabilistic}'s hierarchical sampling model to include contrast, this mechanism would be at odds with a hierarchical Bayesian framework. In contrast, these ideas emerge naturally from the exemplar view: we have not modified any of the core elements of the GCM in defining the PACKER model, we simply aggregated similarity slightly differently. Thus, beyond formally specifying the role of contrast in creative generation, the PACKER account allows us to evaluate the well-understood principles of the exemplar view within the comparatively understudied field of conceptual generation. Nonetheless, we will discuss integrating our approaches in the General Discussion.

@@

\section{Experiment 1}

To begin our investigation, we sought to extend the early evidence obtained from our analysis of the \cite{jern2013probabilistic} data, under a variety of learning conditions and using more standard stimulus materials. We used an artificial stimulus design, two dimensional domain of squares, varying in color and size (see Figure \ref{fig:e1-conditions}, panel A). These dimensions have been used in numerous classification learning studies \cite[e.g.,][]{conaway2016similar,conaway2016generalization,shepard1961learning,nosofsky1994comparing}. Unlike those used in the \cite{jern2013probabilistic} experiments, proximity along these dimensions aligns more directly with perceptual similarity, allowing us to evaluate the role of category contrast in creative generation more precisely. To extend the evidence provided by the \cite{jern2013probabilistic} data, we tested the effects of category contrast after learning one category from a set of qualitatively distinct category structures, as shown in Figure \ref{fig:e1-conditions}.

@@

In this way, we can analyze the results of the first experiment as a conceptual replication of the classic finding that generated categories tend to share distributional properties with known categories in the domain \citep[see][]{jern2013probabilistic,ward1994structured}. From these results, we can predict that, in each condition, participants should generate categories that are distributionally similar to the experimenter-defined category: In the Cluster condition, generated categories should be tightly clustered. In the Row condition, generated categories should vary more along the X-axis than the Y-axis. In XOR condition, generated categories should be widely distributed across both dimensions, and the two dimensions should be positively correlated.

Interestingly, the XOR condition also offers a dissociation between the roles of category contrast and the emulation of distributional structure: widely-distributed, positively-correlated categories would need to lie along the positive diagonal of the space (that is the only place they ``fit''), which is already occupied by the experimenter-defined category. Thus, if contrast plays a role, exemplars in the generated categories of participants in the XOR condition may not be positively correlated -- they may be negatively correlated instead.

@@

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{e1-distanceplots.pgf}
    \caption{Experiment 1 results. {\em Left}: Frequency of generation as a function of distance from members of the experimenter-defined category. {\em Right}: Scatter plot of within-category versus between-category distance in each of the participant-generated categories.}
    \label{fig:e1-distanceplots}
    \end{center}
\end{figure}

Figure \ref{fig:e1-distanceplots} also depicts, for each participant, the average distance of members within the generated category ({\em within-category} distance) against the average distance between members of the generated and experimenter-defined category ({\em between-category} distance). These data also reveal a systematic pattern: the majority of participants generated categories with more between-category distance than within-category distance. That is, members of the generated category tended to be more similar to one another than to members of the experimenter-defined category. To formally evaluate this claim, we conducted t-tests comparing the amount of within- and between- class distance in each condition. All conditions possessed greater between-category distance: Cluster, $t(63) = 11.43$, $p < 0.001$; Row, $t(60) = 13.16$, $p < 0.001$; and XOR, $t(57) = 3.64$, $p < 0.001$. The narrow distribution of between-category distances in the XOR condition reflects the widely distributed nature of the experimenter-defined category: all possible generation candidates are distant from at least one cluster of the experimenter-defined category. These results provide clear evidence of an effect of category contrast: participants prefer to generate categories that are dissimilar to the learned category but maintain some level of internal cohesion.

@@

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{e2-distanceplots.pgf}
    \caption{Experiment 2 results. {\em Left}: Frequency of generation as a function of distance from members of the experimenter-defined category. {\em Right}: Scatter plot of within-category versus between-category distance in each of the participant-generated categories.}
    \label{fig:e2-distanceplots}
    \end{center}
\end{figure}

We began our analysis by again testing for the broad influence of category contrast on generation. As in Experiment 1, we computed the frequency each stimulus was generated as a function of its average distance from members of the experimenter defined category, as well as each participant's average within- and between- category distance. These data, shown in Figure \ref{fig:e2-distanceplots}, yield highly similar results. Stimuli that are more distant from members of the experimenter-defined category were more frequently generated, and the categories in each condition tended to possess more between-category than within-category distance: Bottom, $t(60) = 5.5$, $p < 0.001$; Middle, $t(60) = 2.71$, $p = 0.009$.

We did, however, observe a notable subgroup of participants in each condition who generated categories with more within-category than between-category distance. Upon manual inspection, many of these individuals appear to have assumed a `Corners' strategy, placing exemplars in disparate corners of the space, thus producing much more within-category distance, see Figure \ref{fig:e2-samples} for examples.

@@

In Experiment 2, we replicated the core findings from Experiment 1: stimuli are more likely to be generated if they are distant from exemplars in other categories, and most participants generate categories with more between-category than within-category distance. However, we additionally found that the {\em position} of a previously learned category (rather than its distributional structure) influences the types of categories people generate: participants who learned the `Middle' type were more likely to generate categories spanning the entire Y-axis of the space. Participants who learned the `Bottom' type were less likely to do so as a result of the presence of opposite category exemplars in the lower regions of the space.

@@

Experiments 1 and 2 revealed systematic and strong effects of category contrast on creative generation. In this section, we report the results of simulations with formal models aimed at explaining our observations. Specifically, we present simulations from the PACKER model, as well as a `copy-and-tweak' model (discussed in Section \ref{section:copytweak-packer}), defined as a variant of PACKER with the $\gamma$ parameter constrained to be one. The comparison of these two models serves to highlight the explanatory role of contrast within PACKER's framework: if contrast affords little explanatory advantage, then the two accounts should produce an equally strong account. We also present simulations from an implementation of the hierarchical sampling model proposed by \cite{jern2013probabilistic}, described in-depth in Appendix \ref{ap:hsampling-definition}. The comparison between the hierarchical sampling model and PACKER is meant to emphasize the necessity of contrast and demonstrate it cannot be explained by emulating of distributional structure: whereas PACKER is insensitive to the distributional structure of learned categories (relying only on within- and between-category similarity), the hierarchical sampling model generates categories exclusively on the basis of knowledge of how existing classes are distributed. Our approach in this section is to first broadly evaluate and compare the quality of each model's account to our entire dataset (Experiments 1 and 2 combined), then we more specifically describe the strengths and weakness of each model's account.

@@

Through comparison with the copy-and-tweak model, Figure \ref{fig:packer-loglike} more clearly demonstrates the robustness of the explanatory gains yielded by PACKER's category contrast mechanism. Plotted is PACKER's log-likelihood as a function of $\gamma$ parameter. The model's other parameters ($c$, $\theta$) were set according to copy-and-tweak's best fits from Table \ref{table:global-model-fits}, and thus when $\gamma=1$, the models are equivalent. The figure clearly shows a ``sweet spot'':  a convex region in which PACKER achieves superior fits as a result of changes to $\gamma$. The best fitting values lie well below the value of 1 assumed by the copy-and-tweak model, which demonstrates the robustness of the contrast effect (though note PACKER achieves even better fits when its other parameters are fitted, as in Table \ref{table:global-model-fits}). Notably, however, the copy-and-tweak parameterization ($\gamma=1$) performs better than the `contrast-only' parameterization of the model ($\gamma = 0$). Thus, it is evident that the data is better explained when both within-category similarity and category contrast is considered.

\begin{figure}
    \begin{center}
    \inputpgf{figs/}{packer-loglike.pgf}
    \caption{PACKER's fit as a function of its prioritization of within-category and between-category similarity (using the $\gamma$ parameter). To facilitate comparison, PACKER's other parameters ($c$, $\theta$) were set to the best fitting values obtained for copy-and-tweak in Table \ref{table:global-model-fits}. }
    \label{fig:packer-loglike}
    \end{center}
\end{figure}

@@

In the framework proposed by PACKER, the cluster and corners profiles arise based on different prioritization of within-category similarity versus between-category contrast, and the row and column profiles arise based on the prioritization of each dimension in the computation of similarity. For example, in the cluster profile, there is a high degree of within-category similarity along both dimensions, whereas in the corners profile there is minimal within-category similarity. Thus, PACKER's proposal is that these individual differences arise as a result of different priorities: While the tight cluster configuration can be considered PACKER's `default' mode (as it maximizes within-category similarity), the corners profile can be produced when between-category contrast is put at a higher priority (i.e., $\gamma = 0$).

Likewise, in the row and column profiles, there is a high degree of within-category similarity along one dimension but not the other. These differences likely arise due to a differential focus on one dimension over another, and thus they can be produced by changes to PACKER's attention weights, $w$ (see Equation \ref{eq:similarity}). Traditionally, the attention weights in exemplar models are thought to reflect the diagnostic value of each dimension towards classifying the known category members \citep{nosofsky1984choice,nosofsky1986attention,kruschke1992alcove}, but within a generation context the weights specify the importance of within- and between-category similarity along each dimension. For example, if 100\% of attention is allocated along the X-axis, similarity along the Y-axis no longer influences performance: as a result, PACKER will create categories that are more widely distributed along the Y-axis, as similarity is not counted along that dimension. As a general principle, differentially weighting one dimension will result in the generation of categories that are more widely distributed along the ignored dimension, conforming to a row- or column-like arrangement. See Figure \ref{fig:packer-attention} for a depiction of how attention influences PACKER's performance.

@@

These patterns of performance clearly depict the interdependence between the distributional structure and location of generated concepts. Our results can, in some sense, be interpreted in terms of local minimization of between-category similarity: by distributing the generated category away from members of the experimenter-defined category, participants may increase the degree of between-category distance without drastically altering the degree of within-category similarity.

@@

The creative generation of novel concepts is an intriguing yet understudied topic in cognitive science. While the bulk of prior research on the topic has focused on the classic finding that generated concepts tend to be distributionally similar to known concepts, there has been little work addressing the role of contrast in creative generation: how is it that people are able to create something {\em different} from what is already known? This issue is of fundamental importance: in order to successfully generate something new, it must be different from what is known. We developed a novel, exemplar-based model, PACKER, which formally specifies the role of contrast in generation. The model proposes that categories are represented as exemplars in a multidimensional psychological space, and generation is constrained both by within-category and between-category similarity: exemplars belonging to the same category should be similar to one another, and exemplars belonging to different categories should not be similar to one another.

@@

The success of PACKER yields reveals keys insights about the nature of creative generation. First, by measuring PACKER's fit as a function of its prioritization of within- and between-category similarity, we observed that considering either constraint exclusively results in a relatively low-quality account. Instead, PACKER's best results were obtained when both constraints are considered, indicating that human learners do not generate novel concepts exclusively on the basis of within-category similarity or between class-contrast. This finding mirrors our behavioral results: it is clear that both constraints influence creative generation.

This finding is extended by our observation of the broad interdependence between the distributional structure (feature variance, correlation) and physical instantiation (location within the domain space) of generated categories. In Experiment 2, we found that the unoccupied regions of the domain influenced the distributional structure of categories, and in Experiments 1 and 2 we observed that generated categories assume different distributional structures depending on how they contrast with known categories. This interdependence is straightforwardly explained under PACKER's dual-constraint proposal: after first establishing a location for their generated category, participants distribute exemplars in order to minimize between-category similarity while not drastically decreasing within-category similarity.
